---
title: "FISH 546 Final Writeup"
author: "Aidan Coyle"
date: "3/4/2021"
bibliography: "references.bib"
csl: "ecology.csl"
output:
  html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r citations, echo = FALSE}
# Set up citations
library("knitcitations")
options("citation_format" = "pandoc")
```

## Background - _Hematodinium sp._ and _Chionoecetes bairdi_. 

Before describing the details of my project, we'll examine the system! First, we'll start with the host - the Tanner crab, or _Chionoecetes bairdi_. 

_C. bairdi_ is a true crab native to the North Pacific. It is a widely harvested species, and has quite a substantial commercial, recreational, and subsistence value in Alaska. Within the state, it ranges from southeastern Alaska up to the mid-Bering Sea. Its populations are well-monitored, with annual trawl surveys and pot surveys throughout its Alaskan range. It also has a wide range of parasites, including leeches, chitin-eating bacteria, fungal infections, and the subject of this analysis - the parasitic dinoflagellate _Hematodinium sp._

_Hematodinium sp._ is a broad parasite of many crustaceans, including crabs, amphipods, lobsters, and shrimp. Widely distributed throughout waters north of the equator, it infects many valuable commercial species, including Alaskan snow and Tanner crab (_Chionoecetes spp._). First observed in western Europe in the 1930s, _Hematodinium sp._ was found in Tanner crab of southeast Alaska in 1985 `r citep("10.3354/dao003195")`. Since then, it has become endemic to the region. _Hematodinium_ progresses slowly within _Chionoecetes spp._ with several months between presumed infection and mortality, but is fatal in nearly 100% of cases `r citep("10.3354/dao064253")`. As the infection progresses, _Hematodinium_ proliferates throughout the host, resulting in tissue breakdown and respiratory damage `r citep("10.1007/s42995-020-00061-z")`. The hemolymph becomes an opaque white color, giving the crab a distinct 'cooked' appearance. The crab flesh also becomes bitter and chalky, making the crab unpalatable. This condition is known as Bitter Crab Syndrome (BCS)


## References

``` {r references, include = FALSE}
# write bibliography
write.bibtex(file = "references.bib")

# Fix non-standard characters, otherwise they'll be unicode
# Read in new Bibtex file
fixbib <- readLines("references.bib")
# Remove brackets
fixbib <- gsub(pattern = "<[^>]|>", replace = "", x = fixbib)
# Replace 
fixbib <- gsub(pattern = "+0092", replace = "'", x = fixbib, fixed = TRUE)
writeLines(fixbib, con = "references.bib")
```